Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-31 src/article.tex
second draft

These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. 
While actuators can generate considerable force, it is difficult to achieve the free-motion and resistance of a jack without extensive mechanical intervention such as that carried out by Gillespie \cite{Gillespie1996}. A tactile realism of the harpsichord can already be achieved by the instrument itself. It should first be confirmed if mechanical recreation is sufficient as part of a digital musical instrument before elaborate electromechanical emulation.
% (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?)
, which differ significantly from a piano.

@@

Optical sensors are positioned in front of each jack and their output voltage depends on the amount of light reflected by a sensor surface. In this case, the sensor surface is a grey-scale gradient printed on a vinyl sticker \ref{fig:jack-tags} and applied to the side of each jack. Though the current configuration simply sends MIDI note on and off at passing a threshold, the displacement of all jacks is available continuously.
The staggering of pluck point of the jack means that there are two very distinct tactile `anchors' along the key dip. A combination of string tension, quill voicing and jack staggering contribute to the overall `weight' of the action, with each key differing from the low to the high octaves \cite{Veroli2012}. Such tactile expectations provide fertile opportunities for extending the expressive capabilities of the interface. The current sensor system enables the data recording of traditional use from which it can be identified where those opportunities for new expression lie \cite{McPherson2013-2}.

@@

A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an open source approach for all outputs of the project. The commitment to open sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

@@

\begin{figure}[!b]
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023}
    \label{fig:3key}
\end{figure}

@@

The keyboard was designed to replicate the tactile and aesthetic sensations of playing an antique Italian harpsichord. Traditional materials were used, including walnut for the wrest plank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the 1547 Alessandro Trasuntino harpsichord at the San Colombano Museum. An exception was made for the short octave --- the assigning of common keys in the first octave over a chromatic scale --- typically found in Italian harpsichords: this was replaced by a standard octave layout to more easily accommodate the interaction with commercial sample libraries.

% % Option 2
\begin{figure}
    \centering
     \begin{subfigure}[h]{0.4\linewidth}
        \centering
        \includesvg[inkscapelatex=false, width = \linewidth]{images/log_harp_outline}            
        \caption{logarithmic shape}    
        \label{fig:log-harp}
    \end{subfigure}

@@

The rectangular poplar frame \ref{fig:rect-harp} also deviates from the traditional logarithmic \ref{fig:log-harp} form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string choirs, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations. The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile qualities.

@@

@@

\begin{itemize}
    \item \emph{Non-invasiveness.} No remarkable modifications were allowed on the harpsichord mechanics, particularly on all the visible parts.
    \item \emph{Low Latency.} The time required to process a single iteration of all sensor data should remain under 10 ms, with an upper limit of 25 ms deemed acceptable. Empirical criteria found in previous studies \cite{Jack2016} was used as a guide.
    \item \emph{Reliability:} Sensor data should be dependable and consistent, with interference from anything external to the jack movement being absent or negligible.
    \item \emph{Scalability.} The design needed to scale in both cost and time in order to be adaptable from a 3-key prototype (figure \ref{fig:3key}) up to the 49 keys of the final design.
    % \item \emph{Expandability.} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

@@

The controller board, designed around the Arduino Nano format, was the central hub for processing sensor data and triggering MIDI messages. The in addition to solder terminals for sensor board channels, the controller board contained of:

@@

The Arduino Nano small form factor made fitting the board inside the harpsichord easier and also allowed for testing of multiple chipsets. The 33 BLE variation of the Nano had the additional benefit in its ability to be programmed as BLE MIDI device.  
%Initial iterations of the controller board used 2.54 mm pitch headers for cable connections, but issues with cable disconnections led to the adoption of JST-PH connectors in later designs. Custom cable looms were constructed using adapted crimping tools, ensuring robust connections without compromising accessibility.

@@

Power requirements for the system were estimated at 1.1 A at 5 V, with some fluctuation when the system was first powered on. While the sensors were powered continuously in this iteration, future designs may incorporate power-saving measures, such as dynamic activation of sensors based found in the McPherson piano \cite{McPherson2013}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{src/images/mac-mini.jpg}
    \caption{Components and cabling hidden in the bottom section of the instrument case.}
    \label{fig:mac-mini}
\end{figure}

@@

For the initial version of the exhibition, a harpsichord sample library was used and controlled via Native Instrument's Kontakt. The software was installed on a Mac Mini hosted inside the instrument's case. Holes were drilled at the back of the instrument's frame, away from sight, allow the passage of cables, such as from power supplies, USB connectors and headphone jack \ref{fig:mac-mini}. These components are accessible via locked drawer which is visible in \ref{fig:oratory}. An iPad is used as a monitor through which visitors can adjust parameters of playback such as tuning, voicing and which stops are engaged. The next iteration will use a library of samples from instruments in the collection, which is currently being carried out by the museum.
%A main 12-inch display sits inside a custom-made drawer locked during usage.

\begin{figure}[!hb]
    \centering
    \includegraphics[width=0.8\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=0.8\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}
\subsection{Controller Board}\label{controller-board}

>>>>

These constraints demanded a non-invasive, robust design that was respectful of the historical authenticity of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. 
While actuators can generate considerable force, it is difficult to achieve the free-motion and resistance of a jack without extensive mechanical intervention such as that carried out by Gillespie \cite{Gillespie1996}. A tactile realism of the harpsichord can already be achieved by the instrument itself. It should first be confirmed if mechanical recreation is sufficient as part of a digital musical instrument before elaborate electromechanical emulation.
% (A REFERENCE WOULD BE GOOD HERE, ANY IDEAS?)
% , which differ significantly from a piano.

@@

Optical sensors are positioned in front of each jack and their output voltage depends on the amount of light reflected from a sensor surface. In this case, the sensor surface is a grey-scale gradient printed on a vinyl sticker (Figure \ref{fig:jack-tags}) and applied to the side of each jack. Though the current configuration simply sends MIDI note on and off messages when passing a threshold, the displacement of all jacks is available continuously.
The staggering of the pluck point of the jacks between registers means that there are two very distinct tactile `anchors' along the key dip. A combination of string tension, quill voicing and jack staggering contribute to the overall `weight' of the action, with each key differing from the low to the high octaves \cite{Veroli2012}. Such tactile expectations provide fertile opportunities for extending the expressive capabilities of the interface. The current sensor system enables the data recording of traditional use from which it can be identified where those opportunities for new expression lie \cite{McPherson2013-2}.

@@

A further objective, set forth by the authors, was to ensure accessibility and reproducibility by committing to an open source approach for all outputs of the project. The commitment to open sourcing encompassed all aspects of the system, including hardware schematics, firmware, and calibration data. Cost-effectiveness was also a central consideration. 
% The system was initially developed for a 3-key prototype, shown in Figure \ref{fig:3key}, and successfully scaled to 49 keys without significant increases in cost or complexity. 
Specifically, the system was designed to be easily assembled using resources typically available in a university-managed maker space.

@@

@@

The keyboard was designed to replicate the tactile and aesthetic sensations of playing an antique Italian harpsichord. Traditional materials were used, including walnut for the wrest plank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the 1547 Alessandro Trasuntino harpsichord at the San Colombano Museum. An exception was made for the short octave --- the assigning of common keys in the first octave instead of a chromatic scale --- typically found in Italian harpsichords, which was replaced by a standard octave layout to more easily accommodate the interaction with commercial sample libraries.

% % Option 2
\begin{figure}[!b]
    \centering
     \begin{subfigure}[h]{0.4\linewidth}
        \centering
        \includesvg[inkscapelatex=false, width = \linewidth]{images/log_harp_outline}            
        \caption{logarithmic shape}    
        \label{fig:log-harp}
    \end{subfigure}

@@

The rectangular poplar frame also deviates from the traditional logarithmic \ref{fig:log-harp-comp} form to allow for the installation of the electronic sensors, as visible in Figure \ref{fig:49-key-bottom}, without compromising the visual or tactile authenticity. Two string choirs, crafted from yellow brass wire and anchored with wrought iron pins, were tensioned to replicate authentic plucking resistance. Felt strips were added to dampen vibrations. The result is an interface that combines the mechanical action of keys with synthetic sound generation, preserving a real harpsichord's tactile qualities.

@@

\begin{figure}
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{3-Key Model Harpsichord Mechanism acquired from San Colombano, Bologna, 2023}
    \label{fig:3key}
\end{figure}

@@

\begin{itemize}
    \item \emph{Non-invasiveness.} No remarkable modifications were allowed on the harpsichord mechanics, particularly on all the visible parts.
    \item \emph{Low Latency.} The time required to process a single iteration of all sensor data should remain under 10 ms, with an upper limit of 25 ms deemed acceptable. Empirical criteria found in previous studies \cite{Jack2016} was used as a guide.
    \item \emph{Reliability:} Sensor data should be dependable and consistent, with interference from anything external to the jack movement being absent or negligible.
    \item \emph{Scalability.} The design needed to scale in both cost and time in order to be adaptable from a 3-key prototype (Figure \ref{fig:3key}) up to the 49 keys of the final design.
    % \item \emph{Expandability.} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

@@

The controller board, designed around the Arduino Nano format, was the central hub for processing sensor data and triggering MIDI messages. In addition to solder terminals for the sensor board channels, the controller board contained:

@@

The Arduino Nano small form factor made fitting the board inside the harpsichord easier and also allowed for testing of multiple chipsets. The 33 BLE variation of the Nano had the additional benefit in its ability to be programmed as a BLE MIDI device.  
%Initial iterations of the controller board used 2.54 mm pitch headers for cable connections, but issues with cable disconnections led to the adoption of JST-PH connectors in later designs. Custom cable looms were constructed using adapted crimping tools, ensuring robust connections without compromising accessibility.

@@

Power requirements for the system were estimated at 1.1 A at 5 V, with some fluctuation when the system was first powered on. While the sensors were powered continuously in this iteration, future designs may incorporate power-saving measures, such as dynamic activation of sensors found in the McPherson piano \cite{McPherson2013}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim={0 5cm 0 0},clip]{src/images/mac-mini.jpg}
    \caption{Components and cabling hidden in the bottom section of the instrument case.}
    \label{fig:mac-mini}
\end{figure}

@@

For the initial version of the exhibition, a harpsichord sample library was used and controlled via Native Instrument's Kontakt. The software was installed on a Mac Mini hosted inside the instrument's case. Holes were drilled at the back of the instrument's frame, away from sight, allow the passage of cables, such as from power supplies, USB connectors and headphone jack (Figure \ref{fig:mac-mini}). 
% These components are accessible via locked drawer which is visible in \ref{fig:oratory}. 
An iPad is used as a monitor through which visitors can adjust parameters of playback such as tuning, voicing and which stops are engaged. The next iteration will use a library of samples from instruments in the collection, which is currently being carried out by the museum.
%A main 12-inch display sits inside a custom-made drawer locked during usage.

\begin{figure}[!b]
    \centering
    \includegraphics[width=\linewidth]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=0.8\linewidth]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}
